# `yardstick`

[![Build Status](https://travis-ci.org/topepo/yardstick.svg?branch=master)](https://travis-ci.org/topepo/yardstick)
[![Coverage Status](https://img.shields.io/codecov/c/github/topepo/yardstick/master.svg)](https://codecov.io/github/topepo/yardstick?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/yardstick)](http://cran.r-project.org/web/packages/yardstick)
[![Downloads](http://cranlogs.r-pkg.org/badges/yardstick)](http://cran.rstudio.com/package=yardstick)

```{r load, include = FALSE, message = FALSE, warning = FALSE}
library(yardstick)
library(dplyr)
options(width = 100, digits = 3)
```

`yardstick` is a package to estimate how well models are working using [tidy data](https://www.jstatsoft.org/article/view/v059i10) principals. The package webpage is [`https://topepo.github.io/yardstick/`](https://topepo.github.io/yardstick/) for more information.

For example, suppose you create a classification model and predict a data set. You might have data that look like this:

```{r class-data}
library(yardstick)
library(dplyr)

head(two_class_example)
```

You can use a `dplyr`-like syntax to compute common performance characteristics of the model and get them back in a data frame:

```{r class-metrics}
metrics(two_class_example, truth, predicted)

# or 

two_class_example %>% roc_auc(truth, Class1)
```

[Quasiquotation](http://rlang.tidyverse.org/reference/quasiquotation.html) can also be used:

```{r quasi}
# probability columns:
lvl <- levels(two_class_example$truth)

two_class_example %>% mnLogLoss(truth, !! lvl)
```

## Installation

To install the package:

```{r install, eval = FALSE}
## for development version:
require("devtools")
install_github("topepo/yardstick")
```