---
title: "Metric types"
author: "Davis Vaughan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metric types}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Metric types

There are three main metric types in `yardstick`: class, class probability, and
numeric. Each type of metric has standardized argument syntax, and all metrics 
return the same kind of output (a tibble with 3 columns). This standardization 
allows metrics to easily be grouped together and used with grouped data frames 
for computing on multiple resamples at once. Below are the three types of metrics, along with the
types of the inputs they take.

1) **Class metrics** (hard predictions)

    - `truth` - factor
    
    - `estimate` - factor
  
2) **Class probability metrics** (soft predictions)
    - `truth` - factor
    
    - `estimate / ...` - multiple numeric columns containing class probabilities
  
3) **Numeric metrics**

    - `truth` - numeric
    
    - `estimate` - numeric

## Example

In the following example, the `hpc_cv` data set is used. It contains class
probabilities and class predictions for a linear discriminant analysis fit to
the HPC data set of Kuhn and Johnson (2013). It is fit with 10 fold cross-validation,
and the predictions for all folds are included.

```{r, warning = FALSE, message = FALSE}
library(yardstick)
library(dplyr)
data("hpc_cv")

hpc_cv %>%
  group_by(Resample) %>%
  slice(1:3)
```

1 metric, 1 resample

```{r}
hpc_cv %>%
  filter(Resample == "Fold01") %>%
  accuracy(obs, pred)

```

1 metric, 10 resamples

```{r}
hpc_cv %>%
  group_by(Resample) %>%
  accuracy(obs, pred)
```

2 metrics, 10 resamples

```{r}
class_metrics <- metric_set(accuracy, kap)

hpc_cv %>%
  group_by(Resample) %>%
  class_metrics(obs, estimate = pred)
```

## Metrics

Below is a table of all of the metrics available in `yardstick`, grouped
by type.

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}
library(knitr)
library(dplyr)

yardns <- asNamespace("yardstick")
fns <- lapply(names(yardns), get, envir = yardns)
names(fns) <- names(yardns)

get_metrics <- function(fns, type) {
  where <- vapply(fns, inherits, what = type, FUN.VALUE = logical(1))
  paste0("`", sort(names(fns[where])), "()`")
}

all_metrics <- bind_rows(
  tibble(type = "class", metric = get_metrics(fns, "class_metric")),
  tibble(type = "class prob", metric = get_metrics(fns, "prob_metric")),
  tibble(type = "numeric", metric = get_metrics(fns, "numeric_metric"))
)

kable(all_metrics, format = "html")
```

